---
title: "HW9_TestScores"
Name: Derek Willis/Rob Wells
Date: 10/21/2025
---

name: Madi Eades 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Did a New Reading Program Lead to Better Scores?

The superintendent recently claimed that a new reading program has improved third-grade reading scores across the school district.

Before the program, third-grade students in the district averaged 72.6 points on standardized reading tests with a standard deviation of 4.8 points.

After implementing the program for one semester, you collected scores from 12 randomly selected classrooms:
74, 76, 73, 75, 78, 77, 74, 79, 75, 76, 77, 75

As a journalist, you need to determine: **Is there statistical evidence that reading scores have actually improved?**

## Task 1: Organize your data and initial assessment

Before you can run this codeblock, you will need to fill in a value where it says REPLACE_ME. That value can be found in the introduction.

```{r}
# Known information about reading scores before the new program
prior_mean <- 72.6  # average score
prior_sd <- 4.8     # standard deviation

# Reading scores after implementing the new program (12 classrooms)
new_scores <- c(74, 76, 73, 75, 78, 77, 74, 79, 75, 76, 77, 75) # Replace with the actual scores

# Create a journalist-friendly dataset
score_data <- tibble(
  classroom = paste("Classroom", 1:12),
  reading_score = new_scores
)

# View the data
score_data
```

### Reflection Question 1:  (2 points)
Based on just looking at the score_data dataframe, have test scores improved? How can you tell?
Based solely on the score_data dataframe, it would seem that the class test scores has improved following the implementation of the new reading program. 

## Task 2: Calculate key statistics  (2 points)

Like Task 1, you will need to replace values where it says REPLACE_ME before running any code.


```{r}
# Calculate statistics based on the new reading scores
new_stats <- score_data |> 
  summarise(
    mean = mean(reading_score),
    sd = sd(reading_score),
    n = n()
  )

new_stats
```

### Reflection Question 2: (3 points)
Looking at the mean and standard deviation of the new scores compared to the previous statistics, what initial observations can you make? Provide examples from the data in your answer. What questions might these statistics raise for your reporting?

In the original dataframe, the average test score prior to the reading program was at 72.6, with a standard deviation of 4.8, meaning there is a further spread of the test scores by students around the mean. When the new reading program was introduced, the mean increased to 75.75, with a standard deviation of 1.764. The average test score increased by 3.15 points, suggesting that the new data has higher scores and reflecting that the implementation of the reading program did, in fact, improve the test scores. Additionally, there is a shift in the standard deviation, which also reflects the success of the reading program. The original sd was 4.8, meaning there was a larger spread of test scores around the mean test score, which would bring the average score down. Following the reading program, the standard deviation decreased by almost 3 points to 1.76, which suggests that not only are students scoring better, but they are also scoring similarly to their classmates, reflecting a higher mean value and fewer outliers. 

Questions arisen from the statistics: 
- Why did the scores become more consistent, based on the drop in the standard deviation of the test scores? 
- How do these results compare to district benchmarks, and were they consistent with these outcomes? 
- Was there an overall improvement of test scores for all student groups, or did certain student subsets improve leading to an increase in the average score, while high achievers remained the same across the data? 


## Task 3: Create a column chart  (1 points)

As before, replace any values marked REPLACE_ME based the instructions.


```{r}
# STUDENT TASK: Choose an appropriate fill color for the bars
my_fill_color <- "lightpink" # Replace with a color name like "royalblue", "darkgreen", etc.

# Create a visualization comparing new scores to the previous average
score_data |> 
ggplot(aes(x = classroom, y = reading_score)) +
  geom_col(fill = my_fill_color, alpha = 0.8) +
  geom_hline(yintercept = prior_mean, color = "darkred", size = 1, linetype = "dashed") +
  annotate("text", x = 2, y = prior_mean - 1, 
           label = "Previous Average (72.6)",  hjust = -0.5, vjust= 2, fontface = "bold", color = "darkred") +
  labs(
    title = "Reading Scores After New Program Implementation",
    subtitle = "Horizontal line shows previous district average of 72.6 points",
    x = NULL,
    y = "Reading Test Score",
    caption = "Source: District Assessment Data"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

### Reflection Question 3: (3 points)
Examine the chart you created, and suggest a better title based on the results of the data, not a description. 

"New Reading Program Boosts Test Scores" 

## Task 4: Perform a hypothesis test (2 points)

This is where we formally test the superintendent's claim that reading scores have improved. Fill in the REPLACE_ME values as needed, beginning with your hypotheses.

**Hypotheses:**
Null: The new reading program will not increase students' test scores. 
Alternative: The new reading program will increase students' test scores. 

```{r}
# Set the significance level for your test
alpha_level <- 0.05 # Replace with the appropriate value

# Perform a one-sample t-test
# Since we want to know if scores improved (increased), we use a one-sided test (alternative = "greater")
t_test_result <- t.test(
  score_data$reading_score,
  mu = prior_mean,
  alternative = "greater"
)

# Display the results
t_test_result
```

### Reflection Question 4:  (3 points)
What does the p-value tell you, and what doesn't it tell you? How would you explain these results to a non-technical audience while maintaining accuracy?

From the t-test results, the p-value is 0.00003435. This value is statistically significant since it is far less than 0.05, the alpha value, meaning that the increase in the average test score would be extremely unlikely to happen by chance. 

To a non-technical audience: Student's average reading scores increased from about 72.6 to 75.75, following the introduction of a reading program. 

A statistical test proved that the increase in average test scores for students after the program's implementation was not random. The likelihood of the near 3-point test score increase without the program having any impact is less than 0.05%.  


## Task 5: Interpreting the results for your news story (2 points)

Let's gather all of the important stats we'll need in one place, so we can look at the prior average, the new scores and the results of the t.test, including the confidence interval. Replace any values where it says REPLACE_ME.


```{r}
# Get the p-value
p_value <- t_test_result$p.value

# Calculate the 95% confidence interval
ci <- t.test(score_data$reading_score)$conf.int

# Create a tibble to display the key statistics for your story
story_stats <- tibble(
  `Previous average` = prior_mean,
  `New average` = mean(new_scores),
  `Improvement` = mean(new_scores) - prior_mean,
  `Percent change` = round(((mean(new_scores) - prior_mean) / prior_mean) * 100, 1),
  `p-value` = p_value,
  `Lower bound` = ci[1],
  `Upper bound` = ci[2],
  `Confidence level` = "95%"
)

# Display the key statistics
story_stats
```

## Conclusion

### Reflection Question 5: (4 points)
Based on these statistics, what would be your headline and lead paragraph for this story? Is there evidence to support the superintendent's claim?

Hl: Student test scores see boost after new reading program

Students' test scores increased by 4% following the introduction of a new reading program for the third grade, according to recent test data. 

Yes, there is evidence to back up the superintendent's claim that reading test scores improved after the new reading program for third graders was introduced. First, the average test score jumped by around 3 points after the new reading program, reflecting that after the program was introduced students' test scores improved. Second, the p-value is very close to zero, at 0.00003435, which means that the result of the improvement is statistically significant instead of random chance. Lastly, the t-test tested whether the true mean was greater than 72.6, the original average test score, to show if the scores improved. The lower-bound score fell at 74.63 and the higher-bound interval fell at 76.87, reflecting that the null hypothesis (there was no change at 72.6) can be rejected, since it does not fall within the range. 

### Reflection Question 6:  (1 point)
What metrics or outcomes beyond test scores might be important to track for assessing reading performance?

Other metrics or outcomes beyond test scores that might be important to track for assessing reading performance are students' performance in comprehension assessments, vocabulary growth, reading fluency and possibly teacher observations. Although test scores can be very useful in tracking reading performance for students, it lacks accountability for students who perform poorly on not only reading tests, which can indicate ADHD or test anxiety. Other metrics, such as vocabulary growth and reading fluency, will offer a different perspective on how students perform overall in other aspects of reading performance and how this could possibly influence higher test scores. 
