---
title: "midterm_key_oct_8"
author: "Rob Wells"
date: "2025-10-15"
output: html_document
Madi Eades 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# JOUR405: Statistics for Journalists
## Midterm Exam - Fall 2025

For this exam, you'll analyze several datasets using R and the statistical concepts we've covered in class. Load the tidyverse before beginning, then complete each task. Write your code in the provided blocks and answer the questions in complete sentences.

# Load tidyverse
```{r}
library(tidyverse)
```

## Part 1: Restaurant Health Inspections (25 points)

The first dataset contains restaurant health inspection scores from Montgomery County. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv`

The dataset includes the name of the establishment, the number of points for critical and non-critical areas, the total points, maximum points possible and the compliance score and grade.

## Task 1: Read the data**
```{r}
restaurant_health <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv")
```


## Task 2: Calculate the mean and standard deviation maximum and minimum of compliance scores. Put the results in a new dataframe called "compliance_stats." 
```{r}
compliance_stats <- restaurant_health |> 
  summarize(mean_compliance = mean(compliance_score), 
            sd = sd(compliance_score),
            max_compliance = (max(compliance_score)), 
            min_compliance = (min(compliance_score)),
            median_compliance = median(compliance_score)
  ) 
            
```

## Task 3: Create a histogram of the compliance scores with a vertical line showing the mean. Include a headline. Include a headline and label the x and y axes.
```{r}
restaurant_health |>  
  ggplot() + 
  geom_histogram(aes(x = compliance_score), binwidth = 20) + 
  geom_vline(aes(xintercept = mean(compliance_score)), color = "red", linetype = "dashed", size = 1) + labs(title = "Distribution of Restaurant Health Code Compliance", x = "Compliance Score", y = "Number of Restaurants")
```

## Task 4: Write 2-3 sentences interpreting what the standard deviation and histogram tell us about the distribution of compliance scores. What would be newsworthy about this distribution? What's the story here? 

When you calculate the median of the compliance scores it is 100; therefore, the mean of restaurant's compliance scores is slightly below the median, reflecting that the data is slightly skewed to the left. This is due to the few outliers that scored below 75 points, which in turn brings the average compliance score down and skews the data. However, the mean of 96.30665 suggests that most of the restaurants have high compliance scores, and the distribution of the compliance scores is significantly clustered around this value. There is a tail, skewed to the left, with lower compliance scores, but very few. 


## Part 2: High School Athletics (30 points)

The second dataset shows participation numbers in high school sports across Maryland counties in 2024, broken down by sport and gender. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_hs_participation.csv`

# Task 5: Load the high school participation data
```{r}
hs_athletics <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_hs_participation.csv")
```


#Task 6: Add two columns called total and girls_pct.  
Add two columns called total and girls_pct using mutate(), with the total adding together boys and girls and girls_pct being the percentage of the total represented by girls participants. 
```{r}
hs_athletics |> 
  mutate(
    total = boys + girls, 
    girls_pct = (total - girls) / girls * 10000
  )
```

#Task 7: Calculate correlation between boys and girls participation
```{r}
hs_athletics |> summarize(correlation = cor(boys, girls, method="pearson")) 

```

#Task 8: Create a scatterplot showing this relationship, adding a line of best fit. Include a headline and label the x and y axes.
```{r}
hs_athletics |> 
  ggplot() +
  geom_point(aes(x=boys, y=girls)) +
  geom_smooth(aes(x=boys, y=girls), method="lm") + 
  labs(title = "Girls vs. Boys Participation", x = "Boys", y = "Girls")

```


# Task 9: In 2-3 sentences, explain what the correlation coefficient and scatterplot reveal about gender equity in Maryland high school sports participation. How do you interpret the school districts that are below the line vs those that are above? Which school districts are most worth examining further, and why? (10 points)

Based on the correlation coefficient, 0.982419, there is a high correlation, which can also be reflected by the scatterplot which is a straight upward line with the data close to the line of the scatterplot. The scatterplot shows that majority of the points are below the line; therefore, the two variables do not have a significant relationship. The school districts that would be worth examining further would be Calvert County and Baltimore County Public Schools since these districts are the only districts amongst the data that reflect higher girl participation compared to boy participation in high school athletics. 
## Part 3: Maryland Car Theft Rates (25 points)

Your editor has assigned you a story about car thefts in Maryland and wants you to analyze the data to find out which counties have the highest rates. The fourth dataset contains car theft statistics for Maryland counties in 2023 and population. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv` and complete the following tasks:

#Task 10: Load the car thefts data
```{r}
car_thefts <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv")
```

#Task 11: Using mutate, add a column that calculates the rate of car thefts for each county - you need to choose the per capita rate 
```{r}
theft_rates <- car_thefts |> 
  mutate(
    total_rate = 2023/population * 10000
  )
```

#Task 12: Calculate the median car theft rate and the total number of car thefts statewide. Which counties have rates above the median? 
```{r}
theft_rates |> 
  mutate( 
    median_thefts = median(total_rate),
    total_thefts = 2023
    )
```


#Task 13: Write 2-3 sentences describing what these calculations reveal about the distribution of car thefts in Maryland. What's the lede of a story about your findings? 

The counties that have the highest populations seem to have the lower theft rate values. 

## Part 4: Data Analysis Scenario (20 points)

#Task 14: Essay Question.

You receive a tip that local emergency response times have gotten significantly worse over the past year. You obtain monthly data on response times for police, fire and ambulance calls.

Write 3-4 sentences explaining:
1. What statistical measures would you calculate to verify this claim? (10 points)
2. What visualizations would help readers understand the trends? (5 points)
3. What additional context or data would you need to make this a complete story? (5 points)

**Write your answer here:**

# Excellent Response


### Submission Instructions
- Save your work frequently
- Make sure all code blocks run without errors
- Provide clear explanations for your analytical choices
- Submit both your code and written responses
- Load this test to GitHub and submit the link in Elms

Good luck!
