---
title: "Car Thefts Significance"
author: "Derek Willis/Rob Wells"
date: 10-27-2025
---

name: Madi Eades 


Your editor wants to know if there have been recent changes in the number of car thefts in Prince George's County. Using monthly figures and significance testing, you'll come up with an answer and decide whether there's a story here. As with the previous homework, provide answers in this notebook when prompted and you will have to fill in values where you see REPLACE_ME. Get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# LOAD THE TIDYVERSE
library(tidyverse)
```


### Task 2: Load the Prince George's county crime data and generate car thefts totals by month

```{r}

# Read the data
car_thefts <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/car_thefts_month.csv")  |> 
  arrange(month)  # Sort by date

# Basic inspection of the data
glimpse(car_thefts)

```

### Task 3: Let's make a chart and describe the results.

We'll make a line chart using geom_line() to try and illustrate any trend over time. 

```{r}

car_thefts |> 
ggplot(aes(x = month, y = total)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "darkblue", size = 3) +
  labs(
    title = "Monthly Car Thefts (July 2023 - February 2025)",
    subtitle = "Is there a trend over time?",
    x = "Month",
    y = "Number of Car Thefts"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold")
  )

```


**Q3:** What's your initial reaction to the change in car thefts? Is there a trend?

Initially, it seems there was a sharp decline in car thefts between July of 2024 and January of 2025. Additionally, it seems this decline is a part of a broader trend where there was the highest points of car thefts in July 2023, and though it has fluctuated between the months, it continues to decrease. 


### Task 4: Add the mean, standard error and calculate a margin of error based on a 95% confidence interval

Let's add some stats to that process - mean and standard deviation - and calculate the standard error and 95% confidence intervals. We only have a small number of months, so we'll calculate the margin of error to account for that. 

**INSERT CORRECT VARIABLE AT "REPLACE ME"**

```{r}

theft_summary <- car_thefts |> 
  summarize(
    mean_thefts = mean(total),
    sd_thefts = sd(total),
    months = n(),
    se_thefts = sd_thefts / sqrt(months),
    margin_error = qt(0.975, df = months - 1) * se_thefts,  # Using t-distribution for small sample
    ci_lower = mean_thefts - margin_error,
    ci_upper = mean_thefts + margin_error
  )

# Display the confidence interval
theft_summary
```




### Task 5: Visualize the data

Let's make a better chart showing the changes and the statistics we produced in Task 4. Does this change anything about your answer from Task 3?



```{r}
# Make a better chart
car_thefts |> 
ggplot(aes(x = month, y = total)) +
  geom_line(color = "steelblue", alpha = 0.7) +
  geom_point(size = 2) +
  geom_hline(yintercept = theft_summary$mean_thefts, linetype = "dashed", color = "red", size = 1) +
  geom_ribbon(aes(ymin = theft_summary$ci_lower, ymax = theft_summary$ci_upper), 
              fill = "red", alpha = 0.2) +
  annotate("text", x = min(car_thefts$month), 
           y = theft_summary$mean_thefts + 10, 
           label = paste0("Mean: ", round(theft_summary$mean_thefts, 1)), 
           hjust = 0, color = "red") +
  labs(
    title = "Monthly Car Thefts with 95% Confidence Interval for the Mean",
    x = "Month",
    y = "Number of Car Thefts"
  ) +
  theme_minimal()

```

### Task 6: Describe the confidence interval.

- What does the 95% confidence interval show here? **WRITE ANSWER HERE:** Based on the shaded region, which represents the 95% confidence interval, over half of the points on the graph fall within this range. When points on the graph fall within this region, it reflects the true average number of car thefts over the years. 

- What figure(s) accurately reflects the number of car thefts in a typical month? **WRITE ANSWER HERE:** The figures that accurately reflects the number of car thefts in a typical month is August 2024, which is right on the mean line and within the 95% confidence interval range. This figure accurately represents the average car thefts per month. 

- Propose a headline about the trend in car thefts that is focused on the present and future, but informed by the data you see above. **WRITE ANSWER HERE:** "Car lefts continue to fall after 2023, suggesting sustained decline"



### Task 7: Is the recent trend significant?

It _looks_ like car thefts are decreasing lately. But is that change real? We'll split the monthly data into two halves, and then ask if whether there's a statistically significant difference between these periods. To do that, we'll use a t-test with our alpha at 0.05 and develop a null hypothesis and an alternative hypothesis for this scenario.

Null Hypothesis: there is no real difference between the two periods
Alternative Hypothesis: there is a real difference between the two periods

**INSERT CORRECT VARIABLES AT "REPLACE ME"**

```{r}

# Find the halfway point and add a column placing each month in the first or second halves

midpoint <- ceiling(nrow(car_thefts) / 2)
car_thefts <- car_thefts |> 
  mutate(period = if_else(row_number() <= midpoint, "First Half", "Second Half"))

# Perform t-test comparing the two time periods - this is a two-sided test because we have two periods
t_test_periods <- t.test(total ~ period, data = car_thefts)

# Display results
t_test_periods

# Visualize the comparison
period_summary <- car_thefts |> 
  group_by(period) |> 
  summarize(
    mean_thefts = mean(total),
    sd_thefts = sd(total),
    months = n(),
    se_thefts = sd_thefts / sqrt(months),
    margin_error = qt(0.975, df = months - 1) * se_thefts, # look at how we calculated this above
    ci_lower = mean_thefts - margin_error,
    ci_upper = mean_thefts + margin_error
  )

period_summary |> 
ggplot(aes(x = period, y = mean_thefts, fill = period)) +
  geom_col() +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +
  geom_text(aes(label = round(mean_thefts, 1)), vjust = -0.5, size = 4) +
  labs(
    title = "Comparing Car Thefts: First Half vs. Second Half of Data Period",
    subtitle = paste0("p-value = ", round(t_test_periods$p.value, 3)),
    x = "Time Period",
    y = "Average Number of Car Thefts"
  ) +
  scale_fill_manual(values = c("First Half" = "#66c2a5", "Second Half" = "#fc8d62")) +
  theme_minimal() +
  theme(legend.position = "none")

```

### Task 8: Reflection

Given the above chart, is the difference between the two periods statistically significant? Can we reject the null hypothesis? **WRITE ANSWER HERE:** At first glance, the chart shows that the average car thefts went down from the first half to the second half of the data. In the first half, the average number of car thefts is around 490, and for the second half of the data the average number of car thefts dropped down to 430.8. However, this drop is not statistically significant and the null hypothesis cannot be rejected. According to the t-test, the p-value is 0.179, which is greater than the alpha value 0.05. Therefore, the result of the t-test is not statistically significant, or in other words, the change is not great enough to reject the null hypothesis and suggest there is a real difference in car thefts between the two time periods. 

How would you revise your headline from Task 6, if at all? What would you tell your editor, and what would you expect her to ask you? **WRITE ANSWER HERE:**

"Data shows car thefts are going down, but is this true?" 

If I were reporting on this story, I would tell my editor that the data alone suggests that the average number of car thefts has decreased from 2023 to the beginning of 2025; however, it is essential to understand that this decrease is not necessarily significant, and could simply be a normal variation in the data. I would verify this by telling my editor that the t-test to determine whether there is a statistically significant difference between the first half of the data and the second half of the data gave a p-value of 0.179. I would explain to her that this value is greater than the standard threshold set for the t-test, which was 0.05, and that this means that we cannot say with confidence there is a real difference. 

I would expect my editor to ask me to follow the data to see if this decline will end up being significant, find other datasets regarding car thefts in Prince George's County to verify the results from the data, and how could we frame the story if the decline may not be significant.  



When you are finished, add, commit and push your notebook and submit the URL in ELMS.
