---
title: "HW11: Education Level and Survey Weighting"
author: "Derek Willis/Rob Wells"
date: 10-28-2025
---


Madi Eades

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
```

# Introduction

In this assignment, you'll examine how survey weighting affects our understanding of voting intentions across different education levels, using the data we examined in class. You'll calculate both weighted and unweighted statistics, create visualizations, and reflect on the implications for reporting.

## The Data

The nonvoters dataset contains survey responses about voting intentions and behaviors from a national survey. The survey was conducted prior to an election and includes demographic information like education level.

```{r}
# Load the dataset
nonvoters_data <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/nonvoters_data.csv")

# Take a quick look at the data structure
glimpse(nonvoters_data)
```

### Key Variables

- `weight`: Survey weight assigned to each respondent
- `Q21`: Voting intention (1 = Yes, 2 = No, 3 = Unsure/Undecided)
- `educ`: Education level (College, Some college, High school or less)

## Task 1: Education Distribution

First, let's examine the distribution of education levels in our sample. **Replace "REPLACE_ME"** with the correct variable for education level.

```{r}

education_distribution <- nonvoters_data |>
  count(educ) |>
  mutate(percentage = n / sum(n) * 100)

education_distribution
```

## Task 2: Reflection Question

Why might education levels in survey samples often differ from the general population? What factors might cause certain education groups to be over or underrepresented?

Education levels in survey samples may differ from the general population due to sampling bias, specifically participation bias, in which those with higher levels of education are more likely to respond than those with lower levels. The distribution of education levels amongst the survey participants reflects this; there is a higher percentage of participants with a college-level education in the survey.   

## Task 3: Unweighted Analysis by Education

Now, let's calculate unweighted voting intentions by education level. This is what we would report if we didn't apply any weighting to our sample.

```{r}
# Calculate unweighted voting intentions by education
unweighted_by_education <- nonvoters_data |>
  # Filter out missing values
  filter(!is.na(Q21), Q21 > 0, !is.na(educ)) |>
  # Group by education and response
  group_by(educ, Q21) |>
  # Count responses
  summarize(count = n(), .groups = "drop_last") |>
  # Calculate percentages
  mutate(total = sum(count),
         percentage = count / total * 100) |>
  ungroup()

# Create a more readable format with voting intentions as columns
unweighted_educ_summary <- unweighted_by_education |>
  pivot_wider(
    id_cols = educ,
    names_from = Q21,
    values_from = percentage,
    names_prefix = "pct_"
  ) |>
  rename(
    "Yes (%)" = pct_1,
    "No (%)" = pct_2,
    "Unsure (%)" = pct_3
  )

unweighted_educ_summary
```

## Task 4: Reflection Question

Based just on this unweighted analysis, what headline might you write for a news story about education and voting intentions?

"Higher educated adults more likely to vote, survey data shows" 

## Task 5: Weighted Analysis by Education

Next, let's apply survey weights to see how this changes our results. Instead of just counting responses, we'll sum the weights for each group. **Replace "REPLACE_ME"** with the appropriate weight variable

```{r weighted-by-education}

weighted_by_education <- nonvoters_data |>
  # Filter out missing values
  filter(!is.na(Q21), Q21 > 0, !is.na(educ)) |>
  # Group by education and response
  group_by(educ, Q21) |>
  # Sum the weights instead of counting
  summarize(weighted_count = sum(Q21), .groups = "drop_last") |>
  # Calculate weighted percentages
  mutate(weighted_total = sum(weighted_count),
         weighted_percentage = weighted_count / weighted_total * 100) |>
  ungroup()

# Create a more readable format
weighted_educ_summary <- weighted_by_education |>
  pivot_wider(
    id_cols = educ,
    names_from = Q21,
    values_from = weighted_percentage,
    names_prefix = "pct_"
  ) |>
  rename(
    "Yes (%)" = pct_1,
    "No (%)" = pct_2,
    "Unsure (%)" = pct_3
  )

weighted_educ_summary
```

## Task 6: Reflection Questions

1. How did the percentages change after applying weights? Which education group showed the biggest changes?

When looking at the weighted data by education level, the values vary largely in comparison to the unweighted voting intentions by education level. Some of the education groups that showed the largest changes were the voters who said "yes" and had "high school or less," which was around 74% unweighted and when the voting intentions were weighted it dropped by nearly 20 points to 53%. This indicates that these participants' responses count for less than it would have unweighted. 

2. Why might the weighted results be considered more accurate than the unweighted results?

The weighted results may be more accurate because, when surveys are unweighted, they are not adjusted to represent the population the researchers are sampling from accurately. When survey results are weighted, it corrects for sampling imbalances that result when certain groups within the sample have a higher chance of responding than others. In this example, the weighted results show a more evenly distributed spread of responses and how those respondents intend to vote. For instance, among "some college" respondents, the unweighted values indicate they are highly likely to vote, with nearly 7% indicating they are. However, when the survey results are weighted to adjust for sampling imbalances, the respondents' voting intentions are more evenly distributed, with the "yes" category dropping from 85% to 70% and more respondents in the "unsure" and "no" categories.   

## Task 7: Comparison of Weighted vs. Unweighted Results

Let's create a direct comparison table to see the differences more clearly.

```{r}

comparison <- unweighted_educ_summary |>
  inner_join(weighted_educ_summary, by = "educ", suffix = c("_unweighted", "_weighted")) |>
  mutate(
    # Calculate the differences between weighted and unweighted percentages
    yes_diff = `Yes (%)_weighted` - `Yes (%)_unweighted`,
    no_diff = `No (%)_weighted` - `No (%)_unweighted`,
    unsure_diff = `Unsure (%)_weighted` - `Unsure (%)_unweighted`
  ) |>
  # Select just the columns we want to display
  select(educ, yes_diff, no_diff, unsure_diff) |>
  rename(
    "Education Level" = educ,
    "Yes (% point diff)" = yes_diff,
    "No (% point diff)" = no_diff,
    "Unsure (% point diff)" = unsure_diff
  )

comparison
```

## Task 8: Reflection Question

Which education group shows the largest differences between weighted and unweighted results?

The education groups that show the largest differences between the weighted and unweighted results are the respondents in the "high school or less" category who said "yes," which saw a 20 point difference. This group also saw a 13 point increase between the unweighted and weighted values who said "unsure." Following this education group, the respondents who said "yes" with some college education saw a 14 point decrease between the unweighted and weighted values. 

## Task 9: Visualization

Visualizations can help us see the differences more clearly. Let's create a bar chart comparing weighted and unweighted "Yes" responses by education level. **Replace "REPLACE_ME"** with the correct variable name

```{r}
educ_viz_data <- bind_rows(
  # Unweighted data
  unweighted_by_education |> 
    filter(Q21 == 1) |>  # Only "Yes" responses (Q21=1)
    mutate(Type = "Unweighted") |>
    select(Type, educ, percentage),
  
  # Weighted data - 
  weighted_by_education |> 
    filter(Q21 == 1) |>  # Only "Yes" responses
    mutate(
      Type = "Weighted",
      percentage = weighted_percentage
    ) |>
    select(Type, educ, percentage)
)

# Create a grouped bar chart
ggplot(educ_viz_data, 
       aes(x = educ, y = percentage, fill = Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = sprintf("%.1f%%", percentage)), 
            position = position_dodge(width = 0.9), vjust = -0.5) +
  labs(
    title = "Weighted vs. Unweighted 'Yes' Responses by Education",
    subtitle = "Q21: Do you plan to vote in the November election?",
    y = "Percentage (%)",
    x = "Education Level"
  ) +
  scale_fill_manual(values = c("Unweighted" = "#619CFF", "Weighted" = "#F8766D")) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```


## Task 10: Final Reflection Questions

1. Why is it important for journalists to understand survey weighting when reporting on polls?

It is important for journalists to understand survey weighting when reporting on polls because it can drastically change the summative results of the survey. Without looking at the weighted results, the data could likely be skewed to be overrepresentative of one subsample of the sample compared to another group within the sample, which can lead to misleading and inaccurate reporting. 

2. How might the differences between weighted and unweighted results affect how you would report on this data?

The weighted results would only change slightly for how I owuld report on this data. First, it is important to note that the college educated respondents are still have the highest percentage under "yes" compared to the other groups. Second, it would also be important to report the key difference between the unweighted and weighted results. For example, the unweighted results show far more respondents are likely to vote who have some college education or high school or less compared to the weighted results for these categories. The weighted results suggest there are far more people that are unsure about voting or will not vote at all compared to the unweighted results.  

3. What additional information would you want to know about how the weights were calculated before using this data in a news story?
Some additional information I would like to know about how the weights were calculated before using this data in a news story would be the sample size and how the researchers chose their participants for the survey. 
