---
title: "HW12 car theft anova"
author: "Your Name"
---

Madi Eades


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
```

## Introduction

This analysis examines the impact of a significant spike in car theft rates in 2023 in Prince George's County, and whether that spike has led to a sustained higher rate of car theft in the county. We'll use Analysis of Variance (ANOVA) to determine if theft rates differ significantly between pre-spike (before 2023), spike (2023), and post-spike (2024-present) periods.

## Data Preparation and Exploration

### Loading Libraries and Data

```{r load-libraries}
# Load required libraries
library(tidyverse)
library(lubridate)
```

```{r load-data}
# Import data
theft_data <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/car_thefts_month17_25.csv")

# Look at the first few rows
head(theft_data)

# Get a summary of the data
summary(theft_data)
```


### Get the data how we want it and calculate the rates

**Task 1** Replace REPLACE_ME with the variables you need to calculate the theft_rate (2 points)

```{r process-data}
# Calculate theft rate and create time period variable
theft_data <- theft_data |>
  mutate(
    date = as.Date(month),
    year = year(date),
    month_number = month(date),
    theft_rate = (total / population) * 100000,
    time_period = case_when(
      year < 2023 ~ "Pre-spike",
      year == 2023 ~ "Spike year",
      year > 2023 ~ "Post-spike"
    )
  )

# Create summary by time period
period_summary <- theft_data |>
  group_by(time_period) |>
  summarize(
    mean_thefts = mean(total),
    mean_rate = mean(theft_rate),
    median_rate = median(theft_rate),
    sd_rate = sd(theft_rate),
    n = n(),
    .groups = "drop"
  )

# Display the summary
period_summary
```

**Task 2** Describe the results of calculating the rates for each period. What do they suggest? (3 points)

**WRITE YOUR ANSWER HERE**: The median rate nearly tripled from the 22.2 in the pre-spike to 58.1 in the spike year but the sample sizes varied a great deal, with 71 observations in pre-spike versus 12 in the spike year. The mean rate drops by nearly 15 between the spike year, with 12 reported cases, and the post-spike year. with 14 reported cases. The mean rate decreases between these two time periods; however, the reported cases increased, which suggests that there was an increase in the population that influenced the decrease. Additionally, the mean rate almost doubled from the 23.2 in the pre-spike to 44 in the post-spike, with drastically varying sample sizes, with 71 cases in the pre-spike and 12 cases in the post-spike, which could suggest a subsection of the county is experiencing higher rates of car thefts compared to other parts of the county.  

## Visualizing the Spike Pattern

### Time Series Plot

```{r time-series}
# Time series plot showing the spike pattern
ggplot(theft_data, aes(x = date, y = theft_rate)) +
  geom_line() +
  geom_smooth(method = "loess", span = 0.2) +
  geom_vline(xintercept = as.Date("2023-01-01"), linetype = "dashed", color = "red") +
  geom_vline(xintercept = as.Date("2024-01-01"), linetype = "dashed", color = "red") +
  annotate("text", x = as.Date("2023-07-01"), y = max(theft_data$theft_rate) * 0.9, 
           label = "Spike Year", color = "red") +
  labs(title = "Monthly Car Theft Rates (2017-2025) with Spike Highlighted",
       x = "Date", y = "Thefts per 100,000 Population") +
  theme_minimal()
```

**Task 3** Describe the output of this chart and then give it a better, more active title that provides the lede (3 points)

The chart provides evidence that car thefts indeed spiked during 2023 and began to fall shortly, while remaining above the pre-spike levels of car thefts. 

"Car thefts continue to fall after 2023 spike, but remain above pre-spike numbers"

Data analyzing monthly car theft rates suggests thefts in Prince George's County reached unprecedented numbers in 2023 before sharply decreasing the following year. Yet, the present numbers remain above the the pre-spike levels. 


Now let's take a look at average theft rates.

### Yearly Average Theft Rates

```{r yearly-means}
# Calculate yearly means
yearly_means <- theft_data |>
  group_by(year) |>
  summarize(
    mean_rate = mean(theft_rate),
    .groups = "drop"
  )

# Plot yearly means
ggplot(yearly_means, aes(x = year, y = mean_rate)) +
  geom_line(size = 1) +
  geom_point(size = 3, color = "blue") +
  geom_point(data = yearly_means |> filter(year == 2023), 
             aes(x = year, y = mean_rate), size = 4, color = "red") +
  labs(title = "Average Annual Car Theft Rates",
       x = "Year", y = "Average Thefts per 100,000 Population") +
  theme_minimal() +
  scale_x_continuous(breaks = 2017:2025)
```

## One-way ANOVA Analysis

### Hypothesis Formulation

For our ANOVA test, we'll compare the mean theft rates between the three time periods. First we need to create our hypotheses.

**Task 4** Fill in the hypotheses (2 points)

- **Null Hypothesis (H0)**:   The 2023 car thefts spike has not caused a sustained higher rate of car thefts compared to pre-spike values.

- **Alternative Hypothesis (H1)**: The spike in car thefts in 2023 has caused a sustained higher rate of car thefts compared to pre-spike values. 

### One-way ANOVA Test

```{r anova-test}
# Perform one-way ANOVA on theft rates by time period
period_anova <- aov(theft_rate ~ time_period, data = theft_data)
summary(period_anova)
```

**Task 5** Describe the result: are there real differences in the average rates between the time periods? Is this result significant? (5 points)

First, the F-value is high which suggests the variation between the three time periods is far larger than the variation within the groups themselves. Second, the ANOVA test has a p-value of <2e-16, which is a very low value, far below the significance thresholds, and specifically far less than the standard 0.05 threshold. With this in mind, it can be inferred that the null hypothesis can be rejected that the spike of car thefts in 2023 have not caused a sustained higher rate of car thefts, and, therefore, the alternative hypothesis can be accpeted. 

## Compare the differences between specific periods

Run our post-anova Tukey test and examine the results. 

```{r posthoc}
# Tukey's HSD test
tukey_results <- TukeyHSD(period_anova)
print(tukey_results)
```


```{r period-means}
# Visualize time period means with confidence intervals
period_rates <- theft_data |>
  group_by(time_period) |>
  summarize(
    mean_rate = mean(theft_rate),
    se = sd(theft_rate) / sqrt(n()),
    .groups = "drop"
  ) |>
  mutate(
    lower_ci = mean_rate - qt(0.975, n() - 1) * se,
    upper_ci = mean_rate + qt(0.975, n() - 1) * se
  )

# Ensure time period is ordered correctly
period_rates$time_period <- factor(period_rates$time_period, 
                                   levels = c("Pre-spike", "Spike year", "Post-spike"))

ggplot(period_rates, aes(x = time_period, y = mean_rate, fill = time_period)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = lower_ci, ymax = upper_ci), width = 0.2) +
  labs(title = "Car Theft Rates by Time Period with 95% Confidence Intervals",
       x = "", y = "Mean Theft Rate per 100,000") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Conclusion

This analysis has examined whether there are significant differences in car theft rates between the pre-spike period (before 2023), the spike year (2023), and the post-spike period (after 2023).

**Task 6** Describe the results of the Tukey's test and chart showing confidence intervals. Is there a significant difference between the three periods? In particular, is the difference between the pre-spike and post-spike periods real and significant? Write a headline and lede reporting these findings, focusing on the pre- and post-spike periods. (5 points)

According to Tukey's test, there is a -20 difference, which means the post-spike car theft values are 20 higher than the pre-spike car thefts rate. This supoorts the alternative hypothesis which claims the rate of car thefts has been sustained at a higher rate in the post-spike period compared to the pre-strike period. Furthermore, the chart reflects that following the spike in 2023, the rate of car thefts were maintained at a higher rate compared to the pre-strike rate. The chart also shows the confidence intervals for all three periods, in which between the periods, none of the intervals overlap, suggesting the results of the analysis is statistically significant. More specifically, the pre-spike and the post-spike confidence intervals do not overlap, and because of this we can confidently say the results support the alternative hypothesis. 

"Despite sharp fall after 2023 spike, car thefts remain higher than 2022 numbers in Prince George's County"

Car thefts in Prince George's County reached unprecedented levels in 2023, but quickly declined the next year. However, these values remain significantly higher than the rate of car thefts before the spike. 