---
title: "Multiple Regression Analysis: Maryland Voter Turnout"
name: Madi Eades
date: 12-07-2025
---

# Introduction

Multiple regression analysis is a statistical method used to understand the relationship between one dependent variable and two or more independent variables. This analysis is particularly useful in journalism for exploring complex relationships in civic data, such as how different demographic and socioeconomic factors collectively impact voter turnout. By understanding these relationships, journalists can provide deeper insights into civic participation patterns and identify potential barriers to voting.

```{r setup}
library(tidyverse)
#install.packages('corrplot')
library(corrplot)
```

## Load the Data

The first step in our analysis is to load the data. Here we read the data from a CSV file, which contains information about voter turnout and various demographic factors across Maryland counties. This dataset will be used to build our model that will try to see which factors are most important in predicting voter turnout. The turnout percentage is from the 2024 general election.

```{r}
md_turnout <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_turnout.csv")
head(md_turnout)
```

## Task 1: Exploratory Data Analysis (5 points)

Before building our model, let's explore the data to understand the variables and their relationships. Examine the correlation matrix to see how the independent variables relate to each other and to the dependent variable (voter turnout). Which variables seem to be correlated? Are there any that are not correlated at all? Also look at the multiple scatterplots. Which variables seem to have a linear relationship with the dependent variable? Which do not?

First, the correlation matrix shows that there is a negative correlation between percentage of unemployment and voter turnout, showing a dark red circle, within the -0.6 to -0.8 region. Additionally, the variables that also seem to have a correlation with voter turnout are the percentage of white voters, which has a lighter blue circle suggesting a positive correlation between 0.6 to 0.8. After the percentage of white voters, the following correlated variables with voter turnout are income and median age, within the 0.4 to 0.6 region. 

The variables that seem to have a relationship with voter turnout is voter income, the percentage of white and median age. All of these predictors have positive relationships in predicting voter turnout. On the other hand, the predictor that suggests a negative correlation to voter turnout is the percentage of unemployed voters, which suggests as the percentage of unemployment increases the percentage of voter turnout decreases. 


```{r}
# Summary statistics
summary(md_turnout)

# Create correlation matrix
selected_vars <- md_turnout |> 
  select(turnout, population, income, pct_bachelors_higher, 
         pct_unemployed, median_age, pct_white)
correlation_matrix <- cor(selected_vars)
print(correlation_matrix)

# Visualize correlation matrix
corrplot(correlation_matrix, method = "circle")

# Create scatterplots
md_turnout |>
  pivot_longer(cols = c(population, income, pct_bachelors_higher, 
                       pct_unemployed, median_age, pct_white),
               names_to = "variable", 
               values_to = "value") |>
  ggplot(aes(x = value, y = turnout)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ variable, scales = "free_x") +
  labs(title = "Relationship between predictors and voter turnout",
       y = "Voter turnout (%)")
```

## Task 2: Initial Model Evaluation (5 points)

Next, we construct an initial linear model to analyze how different factors influence voter turnout across Maryland counties. This model helps us quantify the relationships and assess which factors are significant predictors of turnout. 

### Residuals

The residuals of a model are the differences between observed values and the values predicted by the model. For this model:

* **Min and Max** values indicate the range of residuals. 
* **Median** value tells us if the model is generally centered around the actual data points.

### Coefficients

The coefficients tell us the estimated effect of each independent variable on the dependent variable:

* **Population**: Is the coefficient positive or negative? Is it statistically significant?
* **Income**: Does higher income correlate with higher turnout?
* **Education (pct_bachelors_higher)**: How does education level impact voter participation?
* **Unemployment**: Is there a relationship between unemployment and voter turnout?
* **Median Age**: Do counties with older populations have different turnout rates?
* **Race (pct_white)**: Is there a relationship between racial composition and turnout?

### Model Fit

* **Residual Standard Error**: How much does the model typically deviate from actual turnout?
* **Multiple R-squared**: What percentage of the variability in turnout is explained by the model?
* **Adjusted R-squared**: How well does the model fit after accounting for the number of variables?
* **F-statistic**: Is the model as a whole statistically significant?


Examine the output of the model, checking the coefficients, residuals, and overall model fit. How well does this multiple variable model predict turnout? Do any of the variables seem to be insignificant? Is this a useful model, based on what you've seen? Why?

Looking at the residuals, the minimum value is -0.054193 and the maximum value is 0.024262, which are small and suggest that the model can accurately predict the voter turnout against the multiple variables. The median is 0.000494, showing that the data is closely clustered around the zero. 

The variables that seem to be insignificant in predicting voter turnout are education level, unemployment and population, which all have p values that are higher than 0.05. For example, the p-value for education level is 0.83677. 

I think this model can be useful when analyzing the accuracy of initial claims about the relationships between multiple factors that could possibly predict a dependent variable. Knowing the residuals, especially the minimum and maximum values, allows us to examine how accurate the predictions in the model are. Additionally, the p-value of the model is 2.999e-06, which is much lower than 0.05, which suggests the model as a whole is statistically significant. In the data, I originally claimed that there was a negative high correlation between unemployment and voter turnout, which showed a larger circle and linear relationship. Once the model was run, the model suggested the percentage of unemployment was not significant in predicting voter turnout. 

```{r}
# Build initial model with all predictors
initial_model <- lm(turnout ~ population + income + pct_bachelors_higher + 
                   pct_unemployed + median_age + pct_white, data = md_turnout)
summary(initial_model)

```

## Task 3: Model Refinement (5 points)

Create three alternative models by adding or removing variables to see if we can improve the model's performance. Turnout will always be the dependent variable, but you should alter the independent variables by removing or adding some. Then compare the models using adjusted R-squared values. Are any of the models significantly better than the others?

For model 2, I chose percentage of a bachelors degree or higher, aka education level, as the first predictor. This model shows a very small adjusted r-squared value compared to the other models at 0.05044749. This suggests that 5% of the variance in voter turnout is predicted by education level, meaning that education level is not a strong predictor of voter turnout, as expected. 

For model 3, I chose to do income as the predicting variable. The adjusted r-squared value for income level is 0.1656268, which suggests that nearly 17% of the variance in voter turnout can be predicated solely by income level. This model ranked in between model 2 and model 4 on the model comparison graph.  

For model 4, I chose race as the sole predictor of voter turnout in the comparison. The adjusted r-swaured value for racial composition is 0.4569641, or about 46%. This is highly significant compared to the other models in the comparison, reflecting a much higher value than 5 and 17 percent. This suggests that racial composition is a strong indicator in predicting voter turnout. 


```{r}
# Model 2
model2 <- lm(turnout ~ pct_bachelors_higher, data = md_turnout)
summary(model2)

# Model 3
model3 <- lm(turnout ~ income, data = md_turnout)
summary(model3)

# Model 4
model4 <- lm(turnout ~ pct_white, data = md_turnout)
summary(model4)

# Compare models
cat("Model 1 (Full model) - Adjusted R-squared:", summary(initial_model)$adj.r.squared, "\n")
cat("Model 2 - Adjusted R-squared:", summary(model2)$adj.r.squared, "\n")
cat("Model 3 - Adjusted R-squared:", summary(model3)$adj.r.squared, "\n")
cat("Model 4 - Adjusted R-squared:", summary(model4)$adj.r.squared, "\n")

# Create comparison chart
model_names <- c("Full model", "Model 2", "Model 3", "Model 4")
adj_r_squared <- c(summary(initial_model)$adj.r.squared,
                   summary(model2)$adj.r.squared,
                   summary(model3)$adj.r.squared,
                   summary(model4)$adj.r.squared)

model_comparison <- data.frame(Model = model_names, Adjusted_R_squared = adj_r_squared)
ggplot(model_comparison, aes(x = reorder(Model, Adjusted_R_squared), y = Adjusted_R_squared)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Model Comparison by Adjusted R-squared",
       x = "Model")
```


## Task 4: Using the Model (3 points)

Without using the words "multiple regression model", describe how you could use this analysis to cover an election or voter participation story.

Recent data analyzing the 2024 election suggests that racial composition may play the most significant role in predicting voter turnout. 

When analyzing the relationship between voter turnout and voter demographics, including education level and income, the data suggests that race is the strongest predictor of turnout. 

When the results of the data are formed in a manner that simplifies the results, it can guide reporting to look at the gaps in voter turnout, which can be suggested by the racial composition as the highest predictor of turnout. 
