---
title: "HW4 MD CityCrime"
---

In this assignment, we will be working with a dataset of total crimes reported by Maryland cities for two consecutive years. You'll load the data, calculate the violent and property crime rates for each city for both years, and then calculate the percentage change in the crime rates between the two years. You'll also write a short story based on your findings.

### Task 1: Load the Tidyverse and Janitor

You will need to load the tidyverse library and the janitor library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv. You will create a dataframe called `md_cities` that contains the data from this URL.

### Task 2: Load the Data

```{r}
md_cities <- 
  read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv")
```
### Task 3: Calculating Rates

Calculate violent and property crime rates for each city for both years using `mutate()`. You'll create four new columns: `violent_rate_2019`, `property_rate_2019`, `violent_rate_2020`, and `property_rate_2020`. You'll need to figure out the proper unit rate, given that there are varying populations. There's no perfect choice; choose the one that works best for you. Save the results to a new dataframe called `md_cities_rates`.

```{r}
md_cities_rate <- md_cities |> 
  mutate(violent_rate_2019 = violent2019/pop2019 * 100, property_rate_2019 = property2019/pop2019 * 100, violent_rate_2020 = violent2020/pop2020 * 100, property_rate_2020 = property2020/pop2020 * 100) |> 
  select(city, pop2020, pop2019, violent_rate_2019, violent_rate_2020, property_rate_2019, property_rate_2020)
```

### Task 4: Describe the Rates

Examine the new dataframe, `md_cities_rates`, by clicking on it in the Environment tab in RStudio and sorting by the rate columns. Write a couple of sentences about the results, comparing and contrasting the cities with the highest rates (both violent and property) in 2019 and 2020. Do you see any patterns? What does this tell you about crime in Maryland cities? What's the story here?

The data on violent crime rates for both 2019 and 2020 reflect a contrasting image. Unlike I originally expected; cities with smaller populations, like Accident, have high rates of violent crime, comparable to cities with larger populations. In the data, Accident has populations of around 300 people; however, the violent crime rate in 2019 was 32%, and in 2020 it jumped to 59%. These results show a trend among some of the smaller populated cities showing higher rates of violent crime committed than in cities with higher populations. Additionally, the violent crime rates for cities with the highest rates have the highest populations. Baltimore has the highest population and also reports the highest crime rate in 2019; however, Ocean City holds a significant difference over Baltimore in 2020 rates for violent crime, with Ocean City at 239% and Baltimore at almost 160%. Ocean City's crime rate significantly increased in 2020, which could be explained by an increase in tourism during 2020 resulting from post-quarantine. 

The property crime rates for cities in Maryland are higher than the rates of violent crime. For Ocean City, there is a noticeable uptick in property crime from 2019, at 1036.8%, to 2020, with a property crime rate of 1212.8%. For Baltimore, the property crime rate in 2019 was 425%, while the property crime rate in 2020 is almost 315%. Although both city's populations increased between the two years, Ocean City saw a sharp increase in property crime, likely explained by the increased opportunities of burglarizing closed businesses during the pandemic.

Overall, the crime rates that resulted from the data suggest that Maryland has no significant, general trend of crime. With some smaller cities showing unexpectedly high rates of violent crime, and other bigger cities showing higher rates of property crimes. 

For Ocean City, both the violent and property crime rates are relatively high for the city, with comparable rates to Baltimore even though the population in Baltimore is almost seven times the size of Ocean City's. The first story idea that comes to mind is that for decades, Baltimore has been stereotyped as the most violent city in Maryland. The data frame tells a different story, with both violent and property crime steadily decreasing from 2019 to 2020, while other smaller cities actually increasing in both categories. The story should focus on why violent and property crime increase significantly in Ocean City, and Baltimore showing a steady decrease. It would be interesting to understand why Ocean City has higher crime rates than Baltimore, even though both populations increase. This data frame cannot give a full picture of crime within the city, since there are several outlying factors that inhibit gaining a full picture view of crime in Maryland. 

### Task 5: Calculate Percentage Change

Let's add two percentage change columns to your dataframe: one for violent crime rates and one for property crime rates. You'll calculate the percentage change between 2019 and 2020 for each city. Save the results to a new dataframe called `md_cities_change`.

```{r}
md_cities_change <- md_cities_rate |> 
  mutate(
    property_pct_change = (property_rate_2020 - property_rate_2019) / property_rate_2019 * 100, violent_pct_change = (violent_rate_2020 - violent_rate_2019) / violent_rate_2019 * 100
  )

```

### Task 7: Describe Percentage Changes

Look at the `md_cities_change` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change results. What's more interesting here, the patterns or the outliers? What's the story in the data?

ANSWER HERE

### Task 8: Tell the Best Story

Write up a lede and nutgraf of a story with the _best_ story idea you've found in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

ANSWER HERE

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
