---
title: "Regression and Beers"
author: "Wells"
date: "2025-12-02"
output: html_document
---
Madi Eades 

This is another exercise to break down how a simple linear regression works and how to use it in real life.
The topic: beer.

This following exercise comes from Diez, D., Cetinkaya-Rundel, M., & Barr, C. (2024). OpenIntro Statistics (4th ed.). http://www.openintro.org/redirect.php?go=os&referrer=os4_pdf, p 335


Many people believe that gender, weight, drinking habits, and many other factors are much more important in predicting blood alcohol content (BAC) than simply considering the number of drinks a person consumed. Here we examine data from sixteen student volunteers at Ohio
State University who each drank a randomly assigned number of cans of beer. These students were evenly divided between men and women, and they differed in weight and drinking habits. Thirty minutes later, a police officer measured their blood alcohol content (BAC) in grams of alcohol per deciliter of blood.



Load libraries
```{r}
library(tidyverse)
```


Load data
```{r}
bac_data <- read_csv("https://raw.githubusercontent.com/wellsdata/Jour405_student_repo_fall_2025/refs/heads/main/activities/bac%20regression/bac_data.csv")

bac_data

#Student #1 drank 5 beers and has a blood alcohol level of 0.13, etc
```


# Linear Regression
```{r}
# Run the regression
model <- lm(bac ~ beers, data = bac_data)

# Display regression results
cat("\n\nRegression Results:\n")
print(summary(model))
```

 Plot the data and the model line
```{r}
# Create regression equation text
eq_text <- sprintf("BAC = %.4f + %.4f * beers", 
                   coef(model)[1], coef(model)[2])


ggplot(bac_data, aes(x = beers, y = bac)) +
  geom_point(color = "darkblue", size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "red", linewidth = 1) +
  labs(
    x = "Number of Beers",
    y = "Blood Alcohol Content (BAC)",
    title = "BAC vs Number of Beers Consumed"
  ) +
  annotate("text", x = min(bac_data$beers), y = max(bac_data$bac), 
           label = eq_text, hjust = 0, vjust = 1, size = 4) +
  theme_minimal()

```

**Task 1: Interpret the results of this regression. Answer the following questions:**

a) What the predictor or independent variable? What is the response or dependent variable? 
b) What is the regression model asking and what does it tell us?
c) Do the data provide strong evidence that drinking more cans of beer is associated with an increase in blood alcohol? 
d) State the null and alternative hypotheses, report the p-value, and state your conclusion.


**Answer:**
a) The predictor is the number of beers and the response is the blood alcohol level. 
b) The regression model is asking what the relationship is between the amount of beers a student drinks and their blood alcohol level. The model shows us that there is a strong relationship between the two variables, as the number of beers consumed increases, the blood alcohol level increases. 
c) Yes, the data provides strong evidence that drinking more cans, like over 7.5, a student will reach over 0.10.
d) Null hypothesis: As the number of beers a student drinks increases, it will not increase their blood alcohol level. 
Alternative hypothesis: As the number of beers a student drinks increases, their blood alcohol level will also increase. The p-value is 0.0005239, which is far below 0.05, meaning that the results are statistically significant. Overall, the linear regression model showed that there is a strong positive relationship between the amount of beers drank and an increase in blood alcohol level. 


