---
title: "final_project"
output: html_document
date: "2025-11-12"
author: MADI EADES
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Part 1

```{r cars}
library(tidyverse)
library(janitor)
library(dplyr)
```

```{r}
billboard <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vQVR7XE8UGPh9EYKpSLzrNA0kXqcQo3AONREuc6mAAaHAu-8ia01STi_HrHgATfgJ5n9xUAUHPSl9ou/pub?gid=1077686609&single=true&output=csv") |> 
  
clean_names() |> 
mutate(date_new = dmy(date)) |> 
mutate(date_new = if_else(year(date_new) > 2024,  date_new - years(100), date_new)) |> 
select(date_new, everything()) 
```
The data, Billboard Hot 100 Database, that I chose is every song to ever top the Billboard Hot 100 charts from April 1, 1958 to January 11, 2025, compiled by Chris Dalla Riva. Skimming over the dataset, there are several categories, including artist, title, genre, ratings, danceability, etc. Dalla Riva compiled the data as he was writing the book Uncharted Territory: What Numbers tell us about the Biggest Hit Songs and Ourselves https://bio.site/uncharted_territory, which is a data-driven history of hit songs since 1958. The database also includes a data dictionary, which details all of the categories and their definitions. Lastly, the database includes the lyric topics that occurred througout the databases songs. 

For my analysis, the most important columns will be danceability, overall rating, the genre of each song and weeks at number one. For my analysis, I am curious about how danceability may impact the overall rating of a hit song. Additionally, I am not sure what could possibly be missing that I would need to do this analysis as this database is very detailed.

```{r}
billboard |> 
  select(song, overall_rating, danceability) |> 
  arrange(desc(overall_rating), desc(danceability))

```
##danceability
```{r}
danceability_stats <- billboard |>
  summarise(
    mean_danceability   = mean(danceability, na.rm = TRUE),
    max_danceability    = max(danceability, na.rm = TRUE),
    min_danceability    = min(danceability, na.rm = TRUE),
    median_danceability = median(danceability, na.rm = TRUE),
    sd_danceability = sd(danceability, na.rm = TRUE)
  )
```

##Overall Rating: 
```{r}
rating_stats <- billboard |>
  summarise(
    mean_rating   = mean(`overall_rating`, na.rm = TRUE),
    max_rating    = max(`overall_rating`, na.rm = TRUE),
    min_rating   = min(`overall_rating`, na.rm = TRUE),
    median_rating = median(`overall_rating`, na.rm = TRUE),
    sd_rating = sd(`overall_rating`, na.rm = TRUE)
  )
```

##Correlation 
```{r}
billboard |>
  summarise(
    correlation = cor(`overall_rating`, danceability,
                      method = "pearson",
                      use = "complete.obs")
  )
```
The correlation coefficent between Overall Rating and Danceability is 0.024, which is between 0 and 0.3, meaning there is a relationship between the two variables; however, the relationship is weak. If the two variables were to be plotted on a graph, the correlation line would show a slightly upward trend. The correlation coefficient tells us that as one variable goes up, the other slightly increases. 

##scatterplot
```{r}
billboard |> 
  ggplot() +
  geom_point(aes(x=danceability, y=`overall_rating`)) +
  geom_smooth(aes(x=danceability, y=`overall_rating`), method="lm") +
  expand_limits(x = 0, y = 0)
```


## Hypotheses 
Null Hypothesis: Danceability does not influence the overall rating of a song.
Alternative Hypothesis: Danceability does influence the overall rating of a song.

Alternative options to analyze: 

Null hypothesis: The beats per minute and energy scores of a song does not influence weeks the song lasted at number one.   
Alternative hypothesis: The beats per minute and energy of a song does influence weeks the song lasted at number one.

#Engery Stats
```{r}
energy_stats <- billboard |> 
summarise(mean_energy = mean(energy, na.rm = TRUE),
  median_energy = median(energy, na.rm = TRUE), 
  max_energy = max(energy, na.rm = TRUE),
  min_energy = min(energy, na.rm = TRUE), 
  sd_energy = sd(energy, na.rm = TRUE)
)
```

#BPM Stats
```{r}
bpm_stats <- billboard |> 
  summarise(mean_bpm = mean(bpm, na.rm = TRUE),
    median_bpm = median(bpm, na.rm = TRUE),
    max_bpm = max(bpm, na.rm = TRUE),
    min_bpm = min(bpm, na.rm = TRUE), 
    sd_bpm = sd(bpm, na.rm = TRUE)
    )
```

##Questions I have: 

I understand that I need to group the songs by danceability (e.g., high vs. low) and then perform a t-test, but I’m unsure of the best way to implement this in R. What would be the best way to group the danceability scores? 
To test the hypotheses, would it be best to run a t-test on the mean overall rating compared to these two groups? 


##FINAL PROJECT PART TWO 

In this section, I am testing whether or not the race of an artist influences the mean danceability of a song.

Null hypothesis: There is no real difference in the average danceability scores between black artists and white artists. 
Alternative hypothesis: There is a difference in the average danceability score between black artists and white artists, with black artists having more danceable songs. 

To do this, I chose to do a two-sample t-test, comparing the columns artist_white and artist_black. Since the billboard dataset has the artist races in two different columns, with values of 0,1 and 2, I assigned the value of 1 to both groups. Therefore, songs with two or more artists, or not listed at all, were omitted. By assigning the variable of 1 to the dataset, I was able to create two clean groups to run a t-test. Additionally, to understand the amount of artists in each category, I created a table that counted each artist with a assigned value of 1.  

##T-test and grouping 

```{r}
billboard$artist_race <- ifelse(billboard$artist_white == 1, "Artistwhite",
                         ifelse(billboard$artist_black == 1, "Artistblack", NA))

table(billboard$artist_race)

t_test_danceability <- t.test(danceability ~ artist_race, data = billboard)

```
##T-test results and analysis: 

First, from the results of assigning a value of one to each artist, we can see that there are far less black artists (358) than that of white artists (638) in the dataset. To test the if there is a statistically significant difference in danceability scores between songs from white artists and songs from black artists, I used a two sample t-test. The t-test resulted in a p-value of 1.265e-14, or 0.00000000000001265, which is less than the standard p-value cutoff, 0.05. The result of the p-value suggests that there is a statistically significant difference in danceability means between black artists and white artists. Since the p-value is below 0.05, I reject the null hypothesis that there is no real difference in the average danceability score between black artists and white artists, and I accept the alternative hypothesis. 

Additionally, to test whether black artists have more danceable songs than that of white artists I directly compare the mean danceability scores of these two groups from the t-test. The mean danceability score for black artists was much higher, 66.96, than that of white artists, 59.12. These scores signify that black artists, on average, have a more danceable song compared to white artists. 

###Visulization of the results 

```{r}
race_summary <- billboard |>
  group_by(artist_race) |>
  summarize(
    mean_dance = mean(danceability, na.rm = TRUE),
    sd_dance = sd(danceability, na.rm = TRUE),
    songs = n(),
    se_dance = sd_dance / sqrt(songs),
    margin_error = qt(0.975, df = songs - 1) * se_dance,
    ci_lower = mean_dance - margin_error,
    ci_upper = mean_dance + margin_error
  )

race_summary |> 
  ggplot(aes(x = artist_race, y = mean_dance, fill = artist_race)) + 
    geom_col() + 
    geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) + 
  geom_text(aes(label = round(mean_dance, 1)), just = -0.5, size = 4) + 
  labs(
    title = "Average Danceability Score by Artist Race",
    subtitle = paste0("p-value = ", signif(t_test_danceability$p.value, 3)),
    x = 'Artist Race', 
    y = 'Average Danceability') + 
  scale_fill_manual(values = c( 
    "Artistblack" = "lightblue", 
    "Artistwhite" = "lightpink"
     )) + 
  theme_minimal() +
  theme(legend.position = "none")
```
The box plot confirms what the t-test results showed. Black artists have an average danceability score of 67 and white artists have a danceability score of 59.1. According to the box plot, white artists' songs tend to be less danceable than songs from black artists. Additionally, the highest average danceability score went to the group that did not have an indicated race. This strengthens my hypothesis since the graph shows an obvious difference in danceability scores depending on the artist's race. 


##Statistical test #2: Linear Regression for Black Artists 

For my second statistical test, I chose to do a linear regression model to see whether an artist's race (Black v White) can predict danceability scores for a song. I chose to do this test because my hypothesis looks directly at whether an artist's race can predict a more danceable song. 

Null Hypothesis: A song by a Black artist does not predict a higher danceability score
Alternative hypthesis: A song by a Black Artist will predict a higher danceability score

```{r}
model <- lm(`danceability` ~ `artist_black`, data = billboard)
summary(model)
```
##Results

First, the R-squared value is 0.07262, or 7.26%, which is a very low percentage. This percentage suggests that the variation in danceability scores is about 7% explained by the artist's race, showing that race does play a factor in predicting danceability, but only a very small portion of the songs. Second, the p-value is p < 0.001 at 2.2e-16, or 0.00000000000000022, which suggests the results of the model are extremely significant. The regression result on whether a Black artist will predict a higher danceability score is highly statistically significant, which can further prove my alternative hypothesis that A song created by a black artist will predict a higher danceability score.  


##Linear Regression model for white artists 

```{r}
model <- lm(`danceability` ~ `artist_white`, data = billboard)
summary(model)
```
##Results 

Null hypothesis: A song by a Black artist does not predict a higher danceability score
Alternative hypothesis: A song by a white artist will predict a low danceability score

First, the r-squared value is lower than that of the linear model for black artists at 0.01367, or 1.37%. This percentage suggest that only about 1.37% of the variation in danceability scores can be predicted by an artist's race being white. On the other hand, the p-value for the model is 3.484e-05, or 0.00003484, which is far below the thresholds to be statistically significant. This result indicates that if the artist is white it can predict a danceability score.

For both linear regression models, the r-squared values are very low and not close to 100, which indicates that an artist's race cannot be the sole predictor of a song's danceability score. The two-way t-test p-values and the p-values from the linear regression model show that an artist's race is a statistically significant indicator of danceability scores. Tracks featuring Black artists had a higher average than songs featuring white artists; however, it is key to note that an artist's race only predicts a small portion of the danceability score but can still mean it is statistically significant. Therefore, I can reject my null hypotheses and accept my alternative hypotheses. 